Learning to Synthesize Rules for Rule-based Intrusion Detectors

### 2020-03-06 paper/paper.tex
pass on intro

We evaluated \tname{} on a diverse
set of attacks. Results indicate that \Fix{...}
\end{abstract}

@@

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial \nids\ implementations (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar}) exist
today. These systems are very popular in industry to secure local
computer networks given the amount of potential malicious traffic that
exist on the Internet.

This paper focuses on rule-based \nids{}, which are very popular in
industry~\cite{proofpoint-etpro,snort-rule-subscriptions}. A
\emph{rule-based intrusion detector}\footnote{\aka\ signature-based
  intrusion detector.} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:pingscan-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. This
particular rule prescribes a method to capture an Active
Reconnaissance attack for discovering IPS of hosts in a network (see
Section~\ref{sec:active-recon}). Rulesets exist for Snort~\cite{snort}
and Suricata~\cite{suricata}, providing protection against existing
attacks. Unfortunately, attackers actively look for new mechanisms to
exploit systems. For example,
Figure~\ref{fig:distribution-rules-per-month} shows the number of
rules added per month in the year of 2019 to the ``Emerging Threats
Open Ruleset''~\cite{emerging-threats-open}, a popular public data set
of \suri\ rules. We used the ``creation-date'' field associated with
each entry in the ruleset. The data provides some evidence that new
rules are created often.

\begin{wrapfigure}[8in]{r}{0.24\textwidth}
    \vspace{-5ex}  
    \centering
    \scalebox{0.9}{  
      \begin{tikzpicture}
        \begin{axis}[
%            label style={font=\scriptsize},
%            tick label style={font=\scriptsize}
            width = 0.3\textwidth,
            height = 0.2\textwidth,            
            bar width=4pt,
            scaled ticks=false,
            ymajorticks=false,
            ybar stacked,
            enlargelimits=0.09,
            clip=true,
            legend style={at={(0.5,-0.15)},anchor=north,legend columns=-1},          
            xtick=data,
            xticklabel style={rotate=45, font=\footnotesize},
            tick pos=left,
            nodes near coords,
            every node near coord/.append style={font=\footnotesize},
            nodes near coords align={vertical},
            symbolic x coords={Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec},]
            \addplot coordinates {(Jan, 114) (Feb, 155) (Mar, 117) (Apr, 159) (May, 110) (Jun, 231) (Jul, 122) (Aug, 167) (Sep, 107) (Oct, 225) (Nov, 148) (Dec, 127)};
        \end{axis}
      \end{tikzpicture}
    }
    \vspace{-4ex}
    \caption{\label{fig:distribution-rules-per-month}2019 new rules.}
\end{wrapfigure}% <---- don't forget this %

Intuitively, rulesets need to be continuously revised because new
vulnerabilities are discovered and attackers can take advantage of
them. Unfortunately, creating these rules is time-consuming and
error-prone. The language for describing rules is non-trivial (see
Section~\ref{sec:example-suricata-rules}) and testing correctness of a
rule requires data and careful inspection.

@@

This paper proposes \tname{}\footnote{Abbreviation for
  \textbf{Sy}nthesis of Su\textbf{ri}cata R\textbf{u}le\textbf{s}.}, a
technique that uses machine intelligence to automatically synthesize
rules for rule-based \nids. Our goal is to facilitate the creation of
rules and, consequently, help system administrators to harden the
protection of \nids\ to network attacks. \tname{} synthesizes rules to
protect against a given attack from 1) malicious traffic (\ie{},
examples of the attack to protect against), 2) benign traffic, and 3)
trusted rules.

\Fix{
denial-of-service
attack to a server by matching specific conditions about the network
traffic~\cite{understanding-dos}. Relevant properties about the
traffic of interest appear in bold in this rule (see
Section~\ref{sec:suri-metas-coverage}). Deployments of rule-based
\nids\ are restricted to a fixed set of rules defined by the network
system administrator.}

@@

\Fix{summarize how the technique works}

\Fix{summarize results} We believe these results provide initial
evidence on the effectiveness of \tname{}.

@@

@@

\newcommand{\percTopFiveRanking}{66.7\% (=10/15)}

>>>>

We evaluated \tname{} on a diverse
set of attacks. Results indicate that 
\end{abstract}

@@

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar})
implementations exist.

This paper focuses on rule-based \nids{}, which are very popular in
industry~\cite{proofpoint-etpro,snort-rule-subscriptions} given the
amount of potential malicious traffic that exist on the Internet. A
\emph{rule-based intrusion detector}\footnote{\aka\ signature-based
  intrusion detector.} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:pingscan-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. A rule
describes a pattern in the traffic. The observation of that pattern in
the traffic typically indicates an attempt of a mallicious individual
to attack a local network. This particular rule prescribes a method to
capture an Active Reconnaissance attack for discovering accessible
hosts in a network (see Section~\ref{sec:active-recon}). Rulesets
exist for Snort~\cite{snort} and Suricata~\cite{suricata}, providing
protection against existing attacks. Unfortunately, attackers actively
look for new mechanisms to exploit systems through the
network. Figure~\ref{fig:distribution-rules-per-month} shows the
number of rules added per month in the year of 2019 to the ``Emerging
Threats Open Ruleset''~\cite{emerging-threats-open}, a popular public
data set of \suri\ rules. This data provides some evidence that new
rules are created often.

\begin{wrapfigure}[8]{r}{0.24\textwidth}
    \vspace{-3ex}  
    \centering
    \scalebox{0.9}{  
      \begin{tikzpicture}
        \begin{axis}[
            width = 0.3\textwidth,
            height = 0.2\textwidth,            
            bar width=4pt,
            scaled ticks=false,
            ymajorticks=false,
            ybar stacked,
            enlargelimits=0.09,
            clip=true,
            legend style={at={(0.5,-0.15)},anchor=north,legend columns=-1},          
            xtick=data,
            xticklabel style={rotate=45, font=\footnotesize},
            tick pos=left,
            nodes near coords,
            every node near coord/.append style={font=\footnotesize},
            nodes near coords align={vertical},
            symbolic x coords={Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec},]
            \addplot coordinates {(Jan, 114) (Feb, 155) (Mar, 117) (Apr, 159) (May, 110) (Jun, 231) (Jul, 122) (Aug, 167) (Sep, 107) (Oct, 225) (Nov, 148) (Dec, 127)};
        \end{axis}
      \end{tikzpicture}
    }
    \vspace{-4ex}
    \caption{\label{fig:distribution-rules-per-month}2019 new rules.}
\end{wrapfigure}

Intuitively, rulesets need to be continuously revised because the
community finds and advertizes vulnerabilities and attackers can take
advantage of them. Unfortunately, creating these rules is
time-consuming and error-prone. The language for describing rules is
non-trivial (see Section~\ref{sec:example-suricata-rules}) and testing
correctness of a rule requires data and careful inspection.

@@

This paper proposes \tname{}\footnote{Abbreviation for
  \textbf{Sy}nthesis of Su\textbf{ri}cata R\textbf{u}le\textbf{s}.}, a
technique that uses machine intelligence to automatically synthesize
rules for rule-based \nids. Our goal is to facilitate the creation of
rules and, consequently, help system administrators to harden the
protection of \nids\ against network attacks.  \tname{} leverages
different sources of information to synthesize rules for a given
attack. It uses 1) positive examples, \ie{}, scenarios that reproduce
the attack, 2) negative examples in the form of a data set of benign
messages, and 3) a data set of human-written trusted rules. The
positive and negative examples constrain the search space of rules
that can be used to capture the attack. Intuitively, positive examples
indicate what should be captured whereas the negative examples
indicate what should not be captured. It is worth noting that vast
amounts of negative data exist~\cite{tcpreplay,stratosphere-normal},
but positive data is scarce. Consequently, classical machine learning
techniques that require considerable amounts of training data is not
directly applicable in this context. Furthermore, \tname\ uses public
data sets of rules to differentiate rules by the similarity to
human-written rules.

\tname{} takes on input one or more examples of an attack and produces
on output a ranked list of rules for that attack. It uses public data
sets of benign messages~\cite{tcpreplay,stratosphere-normal} and
rulesets~\cite{emerging-threats-open} to create those rules. The
technique is organized as three pipelined components. First,
\tname\ uses the messages that manifest an attack to create a
potentially overspecified seed rule. Then, it uses benign traffic in
to derive from the seed rule a set of plausible rules that isolate the
benign traffic from the malicious traffic. Finally, it uses existing
rules to learn heuristics to rank the plausible rules.

We evaluated \tname\ against 14 attacks of various kinds (see
Table~\ref{table:attacks}). Results are encouraging--- \tname\ ranked
the correct rule among the top five rules in \percTopFiveRanking\ of
the cases and ranked the correct rule among the top ten rules in 11 of
the 14 cases.

%% \Fix{
%% denial-of-service
%% attack to a server by matching specific conditions about the network
%% traffic~\cite{understanding-dos}. Relevant properties about the
%% traffic of interest appear in bold in this rule (see
%% Section~\ref{sec:suri-metas-coverage}). Deployments of rule-based
%% \nids\ are restricted to a fixed set of rules defined by the network
%% system administrator.}

@@

@@

\newcommand{\Contrib}[1]{$\star$#1}
\begin{itemize}[topsep=.2ex,itemsep=.2ex,leftmargin=0.8em]

\item[\Contrib{}]\textbf{Approach.} A novel approach to synthesize
  rules for \nids.

\item[\Contrib{}]\textbf{Evaluation.} An evaluation of our approach on
  attacks of various kinds. Results indicate that \tname\ is effective
  in creating plausible rules and performs very well in ranking the
  correct rule among those generated.

\item[\Contrib{}]\textbf{Implementation.} A public implementation of
  \tname. (Available on request to preserve double-blindness.)

\item[\Contrib{}]\textbf{Data.} A data set of artifacts involved in
  the evaluation of \tname\ (\eg{}, pcap files of example attacks, list
  of plausible rules generated, etc.). The data set is available at
  the following URL~\url{https://zenodo.org/record/3698366}.

\end{itemize}

@@
